Process-based phenotypes: an application to the ontology of cell physiology

### 2011-12-01 paper/drinking.tex

\begin{abstract}
  Phenotypes important for...

  Include behavioral phenotypes, based on behavior ontology

  We identify a problem in representing phenotypes such as drinking
  behavior that...

  provide solution...
\end{abstract}

@@

The large volume and diversity of phenotypes within different species
and across multiple scales and levels of granularity necessitates the
application of flexible strategies for managing and integrating data
so that it becomes amenable to automated comparative analyses. To
integrate biomedical data across heterogeneous information systems,
biomedical ontologies were being developed \cite{Smith2007}. An
ontology is an explicit specification of a conceptualization of a
domain and can be used to make the meaning of terms in a vocabulary
explicit \cite{Gruber1995, Guarino1998}. They play a crucial role in
the annotation of biomedical data and the integration of model
organism databases \cite{go2010, Bada2004, goble}.

@@

Here, we present the foundations for an ontology of process
phenotypes. We present the outline of a theory in which several kinds
of process attributes can be distinguished so that normal and abnormal
physiology of biological systems can be formally characterized. We
apply this theory to cellular processes and create the Cell Phenotype
Ontology (CPO). The CPO is linked to reference ontologies for
qualities, biological processes, functions and cell components, and
its prime application is the unification of phenotypes on the cellular
level across different species as well as for annotation of cellular
phenotypes in domains in which no such ontology exists...

\section{Materials and Methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable amount of
effort and resources in the development and establishment of
ontologies that are becoming increasingly successful as information
management and integration tools in many disparate scientific fields
allowing interoperability and semantic information processing between
diverse biomedical resources and domains.

@@

Based on the axioms and definitions, we can use deduction to infer
statements that logically follow from the axioms.  The process of
automatically deducing sentences from axioms is called {\em automated
  reasoning}. Automated reasoning allows users of an ontology to carry
our key activities: verifying the ontology's consistency, inferring
hidden knowledge and thereby performing powerful queries.  An ontology
is formally inconsistent if there is a statement $\phi$ such that
$\phi$ and its negation $\neg \phi$ can be inferred from the
ontology's axioms.  If an ontology is formally inconsistent, {\em
  every} statement can be inferred from the ontology.

@@

\subsection{PATO and the EQ model}
The Phenotype And Trait Ontology (PATO) was envisaged and designed to
provide a platform for allowing the integration of quantitative and
qualitative phenotype related information across different levels of
granularity of domain knowledge granularity and species
\cite{Gkoutos2005}.  PATO forms an ontology of phenotype qualities
that form basic entities that we can perceive and/or measure such as
colors, sizes, rates etc. One of it's classification axis lies on the
basis of the basic type of entity these qualities inhere to and
distinguishes between qualities of physical objects and qualities of
processes.

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000b},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for the {\tt MP:0009007} term from
the Mammalian Phenotype) it could be used to provide an equivalence
statement between that class and the above PATO-based description.

@@

We use the labels of GO classes to automatically generate class labels
for phenotype classes as well as textual definitions for classes in
the CPO. For example, the label of the class for increased number of
occurrences of {\em Apoptosis} is {\em Increased frequency of
  occurrences of Apoptosis}, and its textual definition states: ``An
increased frequency of occurrences of Apoptosis is a phenotype of
entities in which the number of occurrences of Apoptosis within a
given time period is increased in comparison to a reference process.''

When using the GO and PATO ontologies from 29 Nov 2011, CPO contains
 classes...
Classifies in...

\section{Discussion}
\subsection{Use of CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO BP 'cellular process' as the entity; a further 11\%
reference GO CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration}, {\em decreased
  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
and stands to benefit from the automated synchronization between CPO
and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
body of highly specific, well-supported data to be integrated with
data from other species.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
A domain that would greatly benefit from the development of a CPO is
Systems Microscopy, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
(Lock \& Stromblad, Exp. Cell Res. 2010).

If we consider some of the studies, which have been published in this
field in the last few years (Neumann et al, Nature 464: 722-727;
Schmitz et al, Nature Cell Biology 12: 886-893; Fuchs et al, Mol Sys
Biology 6: 370), the need for a CPO becomes evident.

In all three studies here considered live-cell imaging assays and RNAi
knockdown were used to generate phenotypic profiles that quantify the
cellular response to a given siRNA thus allowing identification of
hundreds of genes involved in diverse biological functions including
cell division, migration and survival.

In each study, several phenotypes were detected and described by the
authors without the use of ontologies (see table), making the
integration between datasets extremely difficult. For example, it is
evident that cell division phenotypes were observed in the three
datasets (e.g. 'Mitotic delay/arrest' and 'Prolonged mitotic exit' or
'Methaphase delay' and 'Methaphase cells') but the overlap between
such phenotypes is unclear.

Data integration is also complicated by the lack of standardization at
the level of data production and processing; all these issues are
currently being address by the different groups involved in the
Systems Microscopy Network of Excellence
(http://www.systemsmicroscopy.eu/) and the first step towards data
integration will be achieved by developing a CPO.

This ontology will be used to integrate phenotypes' definitions across
existing datasets and will then become an integrated part of the data
processing pipeline and used to annotate the data as it gets generated
(Conrad et al, Nature Methods 8: 246-249).

\subsection{Pre- vs. post-composed ontology}
We implemented the CPO using a pattern-based approach to formulating
phenotypes involving processes. The patterns we identify use
pre-existing ontologies, in particular the PATO ontology and the
classification of cellular processes as well as cellular components in
the GO. The result is a large ontology in which classes for phenotypes
are {\em pre-composed}: they are named and defined within an OWL
ontology. However, the large size of the resulting ontology may impair
its utility for data annotation and integration, and software tools
may not always support such very large ontologies. The alternative to
pre-composing all possible phenotype classes using the patterns we
describe is to dynamically generate appropriately defined classes at
the time at which they are being used. To achieve this goal, software
must be developed to support ontology users in applying these patterns
and generate the appropriate class description when required.

% \subsection{Future work}
% Formal, axiomatic system for process qualities, maybe based on PSL,
% FOIS paper 2010

% one example:...

% \subsection{Conclusion}

\section{Acknowledgements}
MAH is supported by Wellcome Trust grant WT090548MA.

>>>>

\begin{abstract}
  The systematic observation of phenotypes has become a crucial tool
  of functional genomics, and several large international projects are
  currently underway to identify and characterize the phenotypes that
  are associated with genotypes in several species. To integrate
  phenotype descriptions within and across species, phenotype
  ontologies have been developed.

  In particular phenotypes on a cellular scale

  \url{http://cell-phenotype.googlecode.com}.
\end{abstract}

@@

The large volume and diversity of phenotypes within different species
and across multiple scales and levels of granularity necessitates the
application of flexible strategies for managing and integrating data
so that it becomes amenable to automated comparative analyses. To
integrate biomedical data across heterogeneous information systems,
biomedical ontologies are being developed \cite{Smith2007}. An
ontology is an explicit specification of a conceptualization of a
domain and can be used to make the meaning of terms in a vocabulary
explicit \cite{Gruber1995, Guarino1998}. They play a crucial role in
the annotation of biomedical data and the integration of model
organism databases \cite{go2010, Bada2004, goble}.

@@

Here, we present the foundations for an ontology of process
phenotypes. We present the outline of a theory in which several kinds
of process attributes can be distinguished so that normal and abnormal
physiology of biological systems can be formally characterized. We
apply this theory to cellular processes and create the Cell Phenotype
Ontology (CPO). The CPO is linked to reference ontologies for
qualities, biological processes, functions and cell components, and
its prime application is the unification of phenotypes on the cellular
level across different species as well as for annotation of cellular
phenotypes in domains in which no such ontology exists.

\section{Materials and Methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable effort and
resources in the development and establishment of ontologies that are
becoming increasingly successful as information management and
integration tools in many disparate scientific fields allowing
interoperability and semantic information processing between diverse
biomedical resources and domains.

@@

Based on the axioms and definitions, we can use deduction to infer
statements that logically follow from the axioms.  The process of
automatically deducing sentences from axioms is called {\em automated
  reasoning}. Automated reasoning allows users of an ontology to carry
out key activities: verifying the ontology's consistency, inferring
hidden knowledge and thereby performing powerful queries.  An ontology
is formally inconsistent if there is a statement $\phi$ such that
$\phi$ and its negation $\neg \phi$ can be inferred from the
ontology's axioms.  If an ontology is formally inconsistent, {\em
  every} statement can be inferred from the ontology.

@@

\subsection{PATO and the EQ model}
PATO was envisaged and designed to provide a platform for allowing the
integration of quantitative and qualitative phenotype related
information across different domains, levels of granularity and
species \cite{Gkoutos2005}.  PATO is an ontology of phenotype
qualities that form basic entities that we can perceive and/or measure
such as colors, sizes, rates etc. One of its classification axes is
based on the basic type of entity to which a qualities belongs, and
PATO distinguishes between qualities of physical objects and qualities
of processes.

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000b},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for example the {\tt MP:0009007}
term from the Mammalian Phenotype) it could be used to provide an
equivalence statement between that class and the above PATO-based
description.

@@

\begin{table*}
  \centering
  \begin{tabular}{|p{7cm}|l|p{4cm}|}
    Neumann et al.\cite{Neumann2010}&Schmitz et
    al.\cite{Schmitz2010}&Fuchs et al.\cite{Fuchs2010}\\
    \hline
    binuclear & normal mitotic exit & actin fiver cells\\
    cell death & prolonged mitotic exit & big cells \\
    cell migration & & bright and large cells phenotype \\
    condensation followed by decondensation without completion of
    mitosis & & bright nuclei\\
    condensation without mitosis/collapse of nucleus & & cells with protrusions\\
    dynamic changes & & elongated cells\\
    failure in decondensation & & elongated cells with protrusions\\
    grape& & high actin ratio cells \\
    increased proliferation& & lamellipodia + high actin ratio cells\\
    large& & lamellipodia cells\\
    large nucleus& & large cells\\
    metaphase alignment problems/including no metaphase& & large nuclei\\
    metaphase delay/arrest& & low eccentricity cells\\
    migration (distance)& & metaphase cells\\
    migration (speed)& & proliferating cells\\
    mitotic delay/arrest& & small cells\\
    nuclei stay close together& & small cells with an enrichment of mitotic cells\\
    polylobed& &\\
    pulsating nuclei& &\\
    segregation problems/chromatin bridges/lagging chromosomes/multiple DNA masses&&\\
    small nucleus&&\\
    strange nuclear shape&&\\
    \hline
  \end{tabular}
  \caption{\label{tbl:studies}The table summarizes cellular phenotype
    terms used in three recent systems microscopy studies.}
\end{table*}

We use the labels of GO classes to automatically generate class labels
for phenotype classes as well as textual definitions for classes in
the CPO. For example, the label of the class for increased number of
occurrences of {\em Apoptosis} is {\em Increased frequency of
  occurrences of Apoptosis}, and its textual definition states that an
increased frequency of occurrences of {\em Apoptosis} is a phenotype
of {\em Regulation of apoptosis} in which the number of occurrences of
{\em Apoptosis} within a given time period is increased in comparison
to a reference process that is considered {\em normal}.

As of November 2011, CPO contains 125,466 classes of which 79,236 are
explicitly defined.  The ELK reasoner \cite{Kazakov2011} is able to
perform a classification of the ontology in under 10 seconds. We make
the ontology and the source code that is used to generate it freely
available on \url{http://cell-phenotype.googlecode.com}.

\section{Discussion}

\subsection{Applications of the CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO-BP's {\em Cellular process} as the entity; a further
11\% reference GO-CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration} and {\em
  decreased occurrence}. FYPO will thus fit neatly under the CPO
umbrella, and stands to benefit from the automated synchronization
between CPO and GO, as well as the integration of cellular phenotypes
across species that the CPO can provide. {\em Schizosaccharomyces
  pombe} annotations to FYPO terms will provide a rich body of highly
specific, well-supported data to be integrated with data from other
species.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A further domain that will greatly benefit from the CPO is {\em
  systems microscopy}, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
\cite{Lock2010}.  If we consider some of the studies, which have been
published in this field in the last few years \cite{Neumann2010,
  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
three studies illustrated in Table \ref{tbl:studies}, live-cell
imaging assays and RNAi knockdown were used to generate phenotypic
profiles that quantify the cellular response to a given siRNA thus
allowing identification of hundreds of genes involved in diverse
biological functions including cell division, migration and survival.
In each study, several phenotypes were detected and described by the
authors without the use of ontologies (see Table \ref{tbl:studies}),
making the integration between datasets extremely difficult. For
example, it is evident that cell division phenotypes were observed in
all three datasets (e.g. {\em Mitotic delay/arrest} and {\em Prolonged
  mitotic exit} or {\em Methaphase delay} and {\em Methaphase cells})
but the overlap between such phenotypes is unclear.

Data integration is also complicated by the lack of standardization at
the level of data production and processing; all these issues are
currently being address by the different groups involved in the
Systems Microscopy Network of Excellence
(http://www.systemsmicroscopy.eu/) and the first step towards data
integration can be achieved by further developing CPO.

This ontology will be used to integrate phenotypes' definitions across
existing datasets and will then become an integrated part of the data
processing pipeline and used to annotate the data as it gets generated
\cite{Conrad2011}.

\subsection{Future research}
We implemented the CPO using a pattern-based approach to formulating
phenotypes involving processes. The patterns we identify are based on
pre-existing ontologies, in particular the PATO ontology and the
classification of cellular processes as well as cellular components in
the GO. The result of our method is a large ontology in which classes
for phenotypes are {\em pre-composed}: they are named and defined
within an OWL ontology. However, the large size of the resulting
ontology may impair its utility for data annotation and integration,
and software tools may not always support such very large
ontologies. The alternative to pre-composing all possible phenotype
classes using the patterns we describe is to dynamically generate
appropriately defined classes at the time at which they are being
used. To achieve this goal, software must be developed to support
ontology users in applying these patterns and generate the appropriate
class description when required.

A further important task is to develop the theory we outlined and
applied for the CPO. In particular, a precise formal characterization
of this theory in terms of axioms will further improve the clarity of
phenotypic descriptions of processes and enable its integration in
well-developed formal ontologies of processes \cite{Herre2006,
  Gruninger2010}.

\section{Acknowledgements}
Funding for RH is provided by the European Commission's 7th
Framework Programme, RICORDO project, grant number 248502. MAH is
supported by Wellcome Trust, grant WT090548MA. HH is supported by the
Institute for Medical Informatics, Statistics and Epidemiology,
University of Leipzig. GVG is funded by NIH grant number R01
HG004838-02.
